### 2020-08-23 thesis/thesis.tex
[thesis] Cite Power and Plotkin Tensors of Comodels
+ TODO section to develop this further
Extend section about coalgebras

    \begin{center}
        $ {\llbracket op \rrbracket}_M : P \times {|I|}^{A} \rightarrow |I| $
    \end{center}

    \noindent
    following now straight from Yoneda Lemma we have:

    \begin{center}
        $ {\llbracket op \rrbracket}_M : {|I|}^{A} \rightarrow |I|^{P} $
    \end{center}

    \noindent
    Let's now take a look on how it dualizes. In \textbf{Set}, we can think about
    $A^{B}$ exponentials as $\mathVar{B-ary}$ coproducts over A, where for each
    argument we select one result. Every arrow in opposite category is reverted,
    thus coproducts becomes products, our morphism is reverted and model becomes
    comodel (\textit{world}). Putting this together, we obtain the following map:

@@

    \noindent
    which is called a \textit{cooperation}. Meaning of that morphism, is that
    based on coalgebra carrier and a parameter, we obtain new state of the
    coalgebra and a value generated by coalgebra.

@@

\noindent
TODO \\
- coalg effects \\
- cohandlers

@@

    \noindent
    where $\Lambda^{0} \subseteq \Lambda$ is a set of $\mathVar{closed~\lambda-terms}$,
    meaning, those terms have no free variables. Let's now define an inductive,
    and coinductive predicate. Relation
    $\Downarrow{} \subseteq \Lambda^{0} \times \Lambda^{0}$ (convergence) for
    call-by-value $\mathVar{\lambda-calculus}$ is defined as follows:

@@

    In fact, libraries for algebraic effects also arose in mainstream languages,
    such as C~\cite{leijen-c} or Python~\cite{python-effect}. In Python,
    effects are implemented through generators~\cite{one-shot}, using built-in
    feature of sending value when doing \textit{yield} operation. Resumptions in
    handlers that are sending value are one-shot and tail-recursive, therefore
    we do not need to handle the coalgebraic part, at the cost of flexibility.

@@

    Work that is the closest related to topic of the thesis, is work done by
    Danel Ahman and Andrej Bauer~\cite{runners-in-action}, as well as their
    implementation of a $\lambda_{coop}$ language~\cite{coop}, which is a
    calculus with runners of coalgebraic effects~\cite{runners-uustalu}
    that ensures linear usage of resources as well as execution of
    code handling finalisation.

\noindent
TODO Consider those\\
- From comodels to coalgebras: State and arrays.\\
- Tensors of comodels and models for operational semantics.\\
- Stateful runners of effectful computations.

>>>>

    \begin{center}
        {$\displaystyle {\llbracket op \rrbracket}_M : P \times {|I|}^{A} \rightarrow |I| $}
    \end{center}

    \noindent
    following now straight from Yoneda Lemma, we have:

    \begin{center}
        {$\displaystyle {\llbracket op \rrbracket}_M : {|I|}^{A} \rightarrow |I|^{P} $}
    \end{center}

    Let's now take a look on how it dualizes. Since we operate in \textbf{Set},
    we can represent $A^{B}$ exponentials as $\mathVar{B-ary}$ products over $A$,
    where for each argument we select one result:

    \begin{center}
        % TODO: a in A under prod
        {$\displaystyle {\llbracket op \rrbracket}_M : \prod_{a \in A}{|I|} \rightarrow \prod_{p \in P}{|I|} $}
    \end{center}

    % TODO: textbf Set^op
    \noindent
    We want now to embed this morphism in $Set^{op}$. Every arrow in opposite
    category is reverted, thus, products become coproducts and our morphism
    is reverted:

    \begin{center}
        {$\displaystyle {\llbracket op \rrbracket}_M : \coprod_{p \in P}{|I|} \rightarrow \coprod_{a \in A}{|I|} $}
    \end{center}

    \noindent
    It turnes out, that this map can be further simplified.

    \begin{lemma}
        In \textbf{Set} category, the following isomorphism holds
        {$\displaystyle \coprod_{b \in B}{A} \cong B \times A $}
    \end{lemma}
    \begin{proof}
        We can think about $\mathVar{B-ary}$ coproducts (tagged unions) over the same set
        such that we have $B$ replicas of $A$ and we select which replica do we want
        to pick. Putting it this way, it immediately follows that it's the same,
        up to isomorphism, as a cartesian product over indexes from $B$ and
        corresponding $A$ sets. For a more detailed and generalized results,
        see~\cite{tensors}.
    \end{proof}

    Model turns into comodel~\cite{comodels}, which we shall also call \textit{world}.
    Putting this together with the above lemma, we obtain the following map:

@@

    \noindent
    which is called a \textit{cooperation}. Meaning of that morphism, is that
    based on coalgebra carrier and a parameter, we obtain new state of the
    coalgebra and a value generated by coalgebra. This state-alike result that
    we have obtained by dualizing models is indeed surprising, and it's
    expected that reader may feel astonished after seeing it for the first time.
    This observation was made in~\cite{tensors}.

@@

    We have now derived a dual concept to algebraic effects and handlers, which
    in the literature are also called comodels and runners~\cite{runners-uustalu, runners-in-action}.
    As we have seen, comodels turned into state-passing (co)operations, which
    in programming langauges, may be a way to model interaction with external
    resources, where based on current configuration and some parameter, we
    obtain new configuration along with it's result.

@@

    \noindent
    where $\Lambda^{0} \subseteq \Lambda$ is a set of $\mathVar{closed~\lambda-terms}$,
    meaning, terms without free variables. Let's now define an inductive,
    and coinductive predicate. Relation
    $\Downarrow{} \subseteq \Lambda^{0} \times \Lambda^{0}$ (convergence) for
    call-by-value $\mathVar{\lambda-calculus}$ is defined as follows:

@@

    In fact, libraries for algebraic effects also arose in mainstream languages,
    such as C~\cite{leijen-c} or Python~\cite{python-effect}. In Python,
    effects are implemented through generators~\cite{one-shot}, using built-in
    feature of sending value when doing \textit{yield} operation. Resumptions in
    handlers that are sending value are one-shot and tail-recursive.

@@

    From theoretical side, research on comodels and it's relation to state
    dates to papers by John Power et.\ al.~\cite{comodels, tensors}, and on
    the runners of comodels to T. Uustalu~\cite{runners-uustalu}.

    Practical part that is closest to topic of the thesis, is work done by
    Danel Ahman and Andrej Bauer~\cite{runners-in-action}, as well as their
    implementation of a $\lambda_{coop}$ language~\cite{coop}, which is a
    calculus with runners of coalgebraic effects that ensures linear usage of
    resources as well as execution of code handling finalisation.
